---
title: "Chapter 3- Data Transformation with dplyr"
author: "Karim Badr"
date: "25/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Prerequisities
```{r}
library(nycflights13)
library(tidyverse)
```

```{r}
flights
```
#Filter rows with filter()
```{r}
filter(flights,month==1,day==1)
```
```{r}
dec25<-filter(flights,month==12,day==25)
dec25
```
```{r}
sqrt(2)^2==2
near(sqrt(2)^2,2)
```
#Logical Operators
```{r}
filter(flights,month==11|month==12)
```
```{r}
filter(flights,month %in% c(11,12))
```
```{r}
filter(flights,!(arr_delay>120|dep_delay>120))
```
```{r}
filter(flights,arr_delay<=120,dep_delay<=120)
```
#Missing Values
```{r}
NA>5
```
```{r}
df<-tibble(x=c(1,NA,3))
filter(df,x>1)
```
```{r}
filter(df,x>1|is.na(x))
```
```{r}
View(flights)
```

#Exercises

```{r}
filter(flights,flights$arr_delay>=120)
```
```{r}
filter(flights,dest=="IAH"|dest=="HOU")
```
```{r}
filter(flights,dest%in% c("IAH","HOU"))
```
```{r}
filter(flights,carrier%in%c("UA","AA","DL"))
```
```{r}
filter(flights,month%in%c(7,8,9))
```
```{r}
filter(flights,dep_delay<=0,arr_delay>120)
```
```{r}
filter(flights,dep_delay>=60,dep_delay-arr_delay>30)
```
```{r}
filter(flights,dep_time <= 600 | dep_time == 2400)
```
```{r}
filter(flights,is.na(dep_time))
```
```{r}
summary(flights)
```

```{r}
NA^0
NA|TRUE
FALSE&NA
NA*0
```
#Arrange rows with arrange()
```{r}
arrange(flights,year,month,day)
```
```{r}
arrange(flights,desc(arr_delay))
```
```{r}
df<-tibble(x=c(5,NA,2))
```

```{r}
arrange(df,x)
```
```{r}
arrange(df,desc(x))
```
```{r}
arrange(df,desc(is.na(x)),x)
```
```{r}
arrange(df,x,desc(is.na(x)))
```

```{r}
arrange(flights, dep_time) %>% 
  tail()
```
```{r}
arrange(flights,desc(is.na(dep_time)),dep_time)
```
```{r}
arrange(flights,desc(dep_delay))
```
```{r}
arrange(flights,dep_delay)
```
```{r}
head(arrange(flights, air_time))
```
```{r}
head(arrange(flights, desc(distance/air_time)))
```
```{r}
head(arrange(flights, desc(distance)))
```
```{r}
head(arrange(flights, distance))
```
#Select Columns with select()
```{r}
select(flights,year,month,day)
```
```{r}
A<-select(flights,year,day,month)
```

```{r}
select(flights,year:day)
```
```{r}
select(flights,-(year:day))
```
```{r}
select(flights,starts_with("d"))
```
```{r}
df1<-tibble(x1=c(1,2),x2=1,x3=4,x5=9,y7=9,x9=8)
```

```{r}
select(df1,num_range("y",1:7))
```
```{r}
rename(flights,tailn909=tailnum)
```
```{r}
select(flights,time_hour,air_time,everything())
```
```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, "dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, 4, 6, 7, 9)
select(flights, all_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
select(flights, any_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, all_of(variables))
variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, !!variables)
```

```{r}
variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, !!variables)
```
```{r}
arrange(flights,!!!variables)
```
```{r}
select(flights,year,year)
```
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
```
```{r}
select(flights,contains("TIME"))
```
```{r}
select(flights, contains("TIME", ignore.case = FALSE))
```
#Add new variables with mutate()
```{r}
flights_sml<-select(flights,year:day,ends_with("delay"),distance,air_time)
```

```{r}
mutate(flights_sml,gain=arr_delay-dep_delay,speed=distance/air_time*60)
```
```{r}
mutate(flights_sml,gain=arr_delay-dep_delay,hours=air_time/60,gain_per_hour=gain/hours)
```
```{r}
transmute(flights_sml,gain=arr_delay-dep_delay,hours=air_time/60,gain_per_hour=gain/hours)
```
#Useful Creation Functions
```{r}
transmute(flights,dep_time,hour=dep_time%/%100,minute=dep_time%%100)
```
```{r}
x<-1:10
lag(x)
lead(x)
```
```{r}
cumsum(x)
```
```{r}
sum(x)
```
```{r}
cummean(x)
```
```{r}
y<-c(1,2,2,NA,3,4)
min_rank(y)
desc(min_rank(y))
min_rank(desc(y))
```
```{r}
row_number(y)
```
```{r}
dense_rank(y)
```
```{r}
percent_rank(y)
```
```{r}
cume_dist(y)
```
```{r}
transmute(flights,dep_time_hr=dep_time%/%100,dep_time_min=dep_time%%100,sched_dep_time_hr=sched_dep_time%/%100,sched_dep_time_min=sched_dep_time%%100)
```
```{r}
transmute(flights,dep_time,dep_time_mins=(dep_time%/%100*60+dep_time%%100)%%1440,sched_dep_time,sched_dep_time_mins=(sched_dep_time%/%100*60+sched_dep_time%%100)%%1440)
```
```{r}
time2mins<-function(x){
  (x%/%100*60+x%%100)%%1440
}
```

```{r}
transmute(flights,dep_time,
  dep_time_mins = time2mins(dep_time),sched_dep_time,
  sched_dep_time_mins = time2mins(sched_dep_time)
)
```
```{r}
(A<-transmute(flights,arr_time,dep_time,air_time,air_time_calc=arr_time-dep_time))
```
```{r}
filter(A,air_time_calc<0)
```
```{r}
(B<-mutate(A,ARR_TIME_MINS=time2mins(arr_time),DEP_TIME_MINS=time2mins(dep_time)))
```
```{r}
B$arr_time-B$dep_time
```
```{r}
B$ARR_TIME_MINS-B$DEP_TIME_MINS
```
```{r}
(C<-select(flights,dep_time,sched_dep_time,dep_delay))
```
```{r}
D<-select(flights,carrier,flight,dep_delay)
D
```
```{r}
D<-mutate(D,dep_delay_rank=min_rank(desc(dep_delay)))
D<-arrange(D,dep_delay_rank)
D
```
```{r}
1:3+1:10
```
Alternatively, we could use the top_n().

flights_delayed3 <- top_n(flights, 10, dep_delay)

#Grouped summaries with summarize()
```{r}
summarise(flights,delay=mean(dep_delay,na.rm=TRUE))
```
```{r}
by_day<-group_by(flights,year,month,day)
summarize(by_day,delay=mean(dep_delay,na.rm = TRUE))
```

#Combining Multiple Operations with the Pipe
```{r}
by_dest<-group_by(flights,dest)
delay<-summarize(by_dest,
                 count=n(),
                 dist=mean(distance,na.rm=T),
                 delay=mean(arr_delay,na.rm = T)
)
delay<-filter(delay,count>20,dest!="HNL")
```
```{r}
ggplot(data = delay,mapping = aes(x=dist,delay))+
  geom_point(aes(size=count),alpha=1/3)+
  geom_smooth(se=FALSE)
```
```{r}
delays<-flights %>% 
  group_by(dest) %>% 
  summarize(count=n(),
            dist=mean(distance,na.rm=T),
            delay=mean(arr_delay,na.rm=T)) %>% 
  filter(count>20,dest!="HNL") 
```
```{r}
delays %>% 
  ggplot(mapping = aes(x=dist,delay))+
  geom_point(aes(size=count),alpha=1/3)+
  geom_smooth(se=FALSE)
```
#Missing Values
```{r}
flights %>% 
  group_by(year,month,day) %>%
```


```{r}
summarize(mean=mean(dep_delay,na.rm=T))
```

```{r}
not_cancelled<-flights %>% 
  filter(!is.na(dep_delay),!is.na(arr_delay)) %>% 
  group_by(year,month) 

not_cancelled %>% 
  summarize(mean_dep_delay=mean(dep_delay),mean_arr_delay=mean(arr_delay)) %>% 
  ggplot()+
  geom_point(mapping=aes(x=month,y=mean_dep_delay),show.legend=TRUE)+
  geom_line(mapping=aes(x=month,y=mean_dep_delay),show.legend=TRUE)+
  geom_point(mapping=aes(x=month,y=mean_arr_delay),color="blue",show.legend=TRUE)+
  geom_line(mapping=aes(x=month,y=mean_arr_delay),color="blue",show.legend=TRUE)
```
```{r}
delays<-not_cancelled %>% 
  group_by(tailnum) %>% 
  summarize(delay=mean(arr_delay))
ggplot(data=delays,mapping=aes(x=delay))+
  geom_freqpoly(bandwidth=10)
```
```{r}
delays<-not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(delay=mean(arr_delay,na.rm = TRUE),n=n())

ggplot(data=delays,mapping=aes(x=n,y=delay))+
  geom_point(alpha=1/10)
```
```{r}
delays %>% 
  filter(n>25) %>% 
  ggplot(mapping=aes(x=n,y=delay))+
  geom_point(alpha=1/10)
```
```{r}
batting<-as_tibble(Lahman::Batting)

batters<-batting %>% 
  group_by(playerID) %>% 
  summarise(
    ba=sum(H,na.rm = TRUE)/sum(AB,na.rm = TRUE),
    ab=sum(AB,na.rm=TRUE)
  )
```
```{r}
batters %>% 
  filter(ab>100) %>% 
  ggplot(mapping=aes(x=ab,y=ba))+
  geom_point()+
  geom_smooth(se=FALSE)
```
```{r}
batters %>% 
  arrange(desc(ba))
```
```{r}
not_cancelled %>% 
  group_by(year,month,day) %>% 
  summarise(
    avg_del_1=mean(arr_delay),
    ave_del_2=mean(arr_delay[arr_delay>0])
  )
```
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(distance_sd=sd(distance)) %>% 
  arrange(desc(distance_sd))
```
```{r}
not_cancelled %>% 
  group_by(month,day,year) %>% 
  summarize(
    first=min(dep_time),
    last=max(dep_time),
    quartile=quantile(dep_time,0.25)
  )
```
```{r}
not_cancelled[1,1:2]
```
```{r}
not_cancelled %>% 
  group_by(year,month,day) %>% 
  summarize(
    first_dep=first(dep_time),
    last_dep=last(dep_time),
    n_th=nth(dep_time,3)
  )
```
```{r}
min_rank(desc(df))
         
```
```{r}
df2<-not_cancelled %>%
  ungroup() %>% 
  select(dep_time)

min_rank(df2)
```
```{r}
not_cancelled %>% 
  group_by(year,month,day) %>% 
  mutate(r=min_rank(desc(dep_time))) %>% 
  filter(r%in%range(r))
```
```{r}
not_cancelled %>%
  ungroup() %>%
  summarize(sum(!is.na(dep_time)))
```
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarize(carriers=n_distinct(carrier)) %>% 
  arrange(desc(carriers))
```
```{r}
not_cancelled %>% 
  ungroup() %>% 
  count(dest)
```
```{r}
not_cancelled %>%
  ungroup() %>% 
  count(tailnum,wt=distance)
```
```{r}
df3<-tibble(x=c("A","B","A","B","C"),y=c(1,3,5,6,2))
df3 %>% 
  count(x,wt=y)
```
```{r}
not_cancelled %>% 
  group_by(year,month,day) %>% 
  summarize(n_early=sum(dep_time<500))
```
```{r}
not_cancelled %>% 
  group_by(year,month,day) %>% 
  summarize(hour_perc=mean(arr_delay>60),n=n(),n1=sum(arr_delay>60))
```
#Grouping by Multiple Variables
```{r}
daily<-group_by(flights,year,month,day)
(per_day<-summarise(daily,flights=n()))
```
```{r}
(per_month<-summarise(per_day,flights=sum(flights)))
```
```{r}
per_year<-summarise(per_month,flight=sum(flights))
per_year
```
```{r}
daily %>% 
  ungroup() %>% 
  summarize(flights=n())
```
#Exercises
#1
#2
```{r}
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))
```

```{r}
not_cancelled %>% 
  count(dest)
```
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(n=n())
```
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  tally()
```
```{r}
not_cancelled %>% 
  count(tailnum,wt=distance)
```
```{r}
not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(n=sum(distance))
```
```{r}
not_cancelled %>% 
  group_by(tailnum) %>% 
  tally(distance)
```
#3
```{r}
Can_Per_Day<-flights %>% 
  mutate(cancelled=is.na(dep_delay)|is.na(arr_delay)) %>% 
  group_by(year,month,day) %>% 
  summarise(
    cancelled_num=sum(cancelled),
    total_flights_num=n()
  )

ggplot(data=Can_Per_Day,mapping=aes(x=total_flights_num,y=cancelled_num))+
  geom_point()
```
```{r}
cancelled_per_day <- 
  flights %>%
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_num = sum(cancelled),
    flights_num = n(),
  )
cancelled_per_day
```


```{r}
(cancelled_and_delays_1<-
  flights %>% 
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year,month,day) %>% 
  summarise(
    prop_canc=mean(cancelled),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
    ))
```
```{r}
ggplot(data=cancelled_and_delays_1)+
  geom_point(mapping = aes(x=prop_canc,y=avg_dep_delay))+
  geom_point(mapping = aes(x=prop_canc,y=avg_arr_delay),color="blue")+
  coord_flip()
```

```{r}
cancelled_and_delays <- 
  flights %>%
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_prop = mean(cancelled),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
  ) 
cancelled_and_delays
```
#5
```{r}
(worst_delays_carrier<-flights %>% 
group_by(carrier) %>% 
summarise(
  dep_delay_mean=mean(dep_delay,na.rm=T),
  arr_delay_mean=mean(arr_delay,na.rm = T)) %>% 
arrange(desc(arr_delay_mean,dep_delay_mean))
)

filter(airlines,carrier=="F9")
```
```{r}
airlines
```
```{r}
worst_delays_carrier %>% 
  arrange(desc(dep_delay_mean))
```
```{r}
flights %>% 
  filter(!is.na(arr_delay)) %>% 
  group_by(origin,carrier,dest) %>% 
  summarise(arr_delay = sum(arr_delay),
    flights= n())%>%
  group_by(origin,dest) %>% 
  mutate(arr_delay_total = sum(arr_delay),
             flights_total=sum(flights)) %>% 
  ungroup() %>% 
  mutate(
    arr_delay_others = (arr_delay_total - arr_delay) /
      (flights_total - flights),
    arr_delay_mean = arr_delay / flights,
    arr_delay_diff = arr_delay_mean - arr_delay_others
  ) %>%
  # remove NaN values (when there is only one carrier)
  filter(is.finite(arr_delay_diff)) %>%  # average over all airports it flies to
  group_by(carrier) %>%
  summarise(arr_delay_diff = mean(arr_delay_diff)) %>%
  arrange(desc(arr_delay_diff))
```
```{r}
flights %>% 
  filter(!is.na(arr_delay)) %>% 
  group_by(origin,carrier,dest) %>% 
  summarise(arr_delay = sum(arr_delay),
    flights= n())%>%
  group_by(origin,dest) %>% 
  mutate(arr_delay_total = sum(arr_delay),
             flights_total=sum(flights)) 
```
```{r}
flights %>%
  count(dest, sort = TRUE)
```
```{r}
(EX57<-flights %>%
  filter(!is.na(dep_delay)) %>% 
  select(year,month,day,tailnum,dep_delay) %>% 
  arrange(tailnum,year,month,day) %>% 
  group_by(tailnum) %>% 
  mutate(cumulative_hr_delays = cumsum(dep_delay > 60)) %>%  
  summarise(total_flights = sum(cumulative_hr_delays < 1))
)
```
```{r}
EX57 %>% 
  filter(tailnum=="N0EGMQ",dep_delay>60)
```
#Grouped Mutates (and Filters)
```{r}
flights_sml %>% 
  group_by(year) %>% 
  filter(rank(desc(arr_delay))<10) %>% 
  arrange(arr_delay)
```
```{r}
A<-arrange(flights_sml,desc(arr_delay)) %>%   
mutate(rank=rank(desc(A$arr_delay)))
```

```{r}
popular_dests<-flights %>% 
  group_by(dest) %>% 
  filter(n()>365)  

ABC<-popular_dests %>% summarize(n=n()) 
min(ABC$n)
```
```{r}
(ABCD<-popular_dests %>% 
  filter(arr_delay>0) %>% 
  mutate(prop_delay=arr_delay/sum(arr_delay)) %>%
  select(dest,arr_delay,prop_delay) %>% 
  arrange(dest)) %>% 
  summarize(sum(arr_delay),sum(prop_delay))
```
```{r}
ABCD %>% 
  summarize(sum(prop_delay))
```

```{r}
ABCD %>% 
  select(dest,arr_delay) %>% 
  filter(dest=="ALB") %>% 
  summarise(n=sum(arr_delay))
```
```{r}
EX1<-tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3))
```

```{r}
EX1 %>% 
  mutate(x_mean = mean(x)) %>% 
  group_by(group) %>% 
  mutate(x_mean1 = mean(x))
```
```{r}
tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3)) %>%
 mutate(lag_x = lag(x),
         lead_x = lead(x)) %>% 
  group_by(group) %>% 
  mutate(lag_x_1=lag(x),lead_x_1=lead(x))
  
```
#Exercises
2
```{r}
flights %>%
  filter(!is.na(tailnum)) %>%
  mutate(on_time = !is.na(arr_time) & (arr_delay <= 0)) %>%
  group_by(tailnum)%>%
  summarise(on_time = mean(on_time), n = n()) %>% 
  filter(min_rank(on_time) == 1)
```
```{r}
flights %>%
  filter(!is.na(tailnum), is.na(arr_time) | !is.na(arr_delay)) %>%
  mutate(on_time = !is.na(arr_time) & (arr_delay <= 0)) %>%
  group_by(tailnum) %>%
  summarise(on_time = mean(on_time), n = n()) %>%
  filter(n >= 20) %>%
  filter(min_rank(on_time) == 1)
```
```{r}
flights %>%
  filter(!is.na(arr_delay)) %>%
  group_by(tailnum) %>%
  summarise(arr_delay = mean(arr_delay), n = n()) %>%
  filter(n >= 20) %>% 
   filter(min_rank(desc(arr_delay)) == 1)
```
```{r}
flights %>% 
  group_by(hour) %>% 
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(arr_delay)
```
```{r}
flights %>% 
  filter(arr_delay>0) %>% 
  group_by(dest) %>%
  mutate(
    arr_delay_total = sum(arr_delay),
    arr_delay_prop = arr_delay / arr_delay_total
  ) %>%
  select(dest, month, day, dep_time, carrier, flight,
         arr_delay, arr_delay_prop) %>%
  arrange(dest, desc(arr_delay_prop))
  
```
```{r}
flights %>%
  filter(arr_delay > 0) %>%
  group_by(dest, origin, carrier, flight) %>%
  summarise(arr_delay = sum(arr_delay)) %>%
  group_by(dest) %>%
  mutate(
    arr_delay_prop = arr_delay / sum(arr_delay)
  ) %>%
  arrange(dest, desc(arr_delay_prop)) %>%
  select(carrier, flight, origin, dest, arr_delay_prop)
```
```{r}
lagged_delays <- flights %>%
  arrange(origin, month, day, dep_time) %>%
  group_by(origin) %>%
  mutate(dep_delay_lag = lag(dep_delay)) %>%
  filter(!is.na(dep_delay), !is.na(dep_delay_lag))
lagged_delays
```
```{r}
lagged_delays %>%
  group_by(dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 1500, by = 120)) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
```
```{r}
lagged_delays %>%
  group_by(origin, dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  facet_wrap(~ origin, ncol=1) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
```
```{r}
standardized_flights <- flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest, origin) %>%
  mutate(
    air_time_mean = mean(air_time),
    air_time_sd = sd(air_time),
    n = n()
  ) %>%
  ungroup() %>%
  mutate(air_time_standard = (air_time - air_time_mean) / (air_time_sd + 1))
ggplot(standardized_flights, aes(x = air_time_standard)) +
  geom_density()
```
```{r}
standardized_flights %>%
  arrange(air_time_standard) %>%
  select(
    carrier, flight, origin, dest, month, day,
    air_time, air_time_mean, air_time_standard
  ) %>%
  head(10) %>%
  print(width = Inf)
```
```{r}
flights %>%
   # find all airports with > 1 carrier
   group_by(dest) %>%
   mutate(n_carriers = n_distinct(carrier)) %>%
   filter(n_carriers > 1) %>%
   # rank carriers by numer of destinations
   group_by(carrier) %>%
   summarize(n_dest = n_distinct(dest)) %>%
   arrange(desc(n_dest))
#> `summarise()` ungrouping output (override with `.groups` argument)
#> # A tibble: 16 x 2
#>   carrier n_dest
#>   <chr>    <int>
#> 1 EV          51
#> 2 9E          48
#> 3 UA          42
#> 4 DL          39
#> 5 B6          35
#> 6 AA          19
#> # … with 10 more rows
```



